---
title: "EX2104: 2021 North Atlantic Stepping Stones"
format:
  docx: 
    reference-doc: 2025_OER_for-quarto.docx
editor: visual
---

```{r}
#| echo: false
#| output: false

#set expedition information to use throughout document
expedition_name <- "EX2104"
expedition_title <- "North Atlantic Stepping Stones"
annotations_date <- "October 21, 2024"
UCH_dives <- 0

#libraries
library(readr)
library(ggplot2)
library(showtext)
library(tidyr)
library(dplyr)
library(forcats)

#add Roboto to showtext package
font_add_google("Roboto", "Roboto")
showtext_auto()

#create OER figure theme to apply throughout the report
OER_theme <- theme_bw() + theme(plot.title = element_text(family = "Roboto", color = "#003087", size = 16),
        axis.title = element_text(family = "Roboto", color = "#003087", size = 12),
        axis.text = element_text(family = "Roboto", size = 10))

#load summary statistics
summary_statistics <- read_csv(paste0("summary_statistics_",expedition_name,".csv"),col_names = TRUE, show_col_types = FALSE)
#load phylum frequency percent
phyla_frequency_percent_all <- read_csv(paste0("phyla_frequency_percent_all_",expedition_name,".csv"),col_names = TRUE)

#calculate total dive number
science_dive_count <- nrow(summary_statistics)
total_dive_count <- science_dive_count + UCH_dives

#calculate percentage of unidentified annotations at each taxonomic level
annotations_taxonomy_percentage <- summary_statistics |> 
  mutate(across(class:species, \(x) x/phylum*100, .names = "{.col}_percent")) |> 
  select("dive_number", class = "class_percent", order = "order_percent", family = "family_percent", genus = "genus_percent", species = "species_percent") |> 
  pivot_longer(class:species) |> 
  rename(taxonomic_level = name,
         annotation_percentage = value) |> 
  mutate(taxonomic_level = as.factor(taxonomic_level),
         dive_number = as.factor(dive_number)) |> 
  mutate(taxonomic_level = fct_relevel(taxonomic_level,
                                       c("class","order",
                                         "family","genus","species")))
annotations_taxonomy_percentage[is.na(annotations_taxonomy_percentage)] = 0

#perform min/max/median calculations across dives for each data type

#ROV bottom time
bottom_min <- round(min(summary_statistics$bottom_time_hours),2)
bottom_max <- round(max(summary_statistics$bottom_time_hours),2)
bottom_median <- round(median(summary_statistics$bottom_time_hours),2)

#Mean depth
depth_min <- format(round(min(summary_statistics$mean_depth),0),big.mark = ",")
depth_max <- format(round(max(summary_statistics$mean_depth),0), big.mark = ",")
depth_median <- format(round(median(summary_statistics$mean_depth),0), big.mark = ",")

#number of geoform/substrate annotations
geo_sub_annotations_min <- format(min(summary_statistics$geoform_or_substrate),big.mark = ",")
geo_sub_annotations_max <- format(max(summary_statistics$geoform_or_substrate),big.mark = ",")
geo_sub_annotations_median <- format(median(summary_statistics$geoform_or_substrate),big.mark = ",")

#number of biological annotations
bio_annotations_min <- format(min(summary_statistics$phylum),big.mark = ",")
bio_annotations_max <- format(max(summary_statistics$phylum),big.mark = ",")
bio_annotations_median <- format(median(summary_statistics$phylum),big.mark = ",")

#percentage annotations flagged for review
percent_flagged_min <- round(min(summary_statistics$percent_flagged),1)
percent_flagged_max <- round(max(summary_statistics$percent_flagged),1)
percent_flagged_median <- round(median(summary_statistics$percent_flagged),1)

#number of unidentified biological annotations
unID_bio_min <- format(min(summary_statistics$Unidentified_Biota),big.mark = ",")
unID_bio_max <- format(max(summary_statistics$Unidentified_Biota),big.mark = ",")
unID_bio_median <- format(median(summary_statistics$Unidentified_Biota),big.mark = ",")

#number of Cnidaria annotations
cnidaria_min <- format(min(summary_statistics$Cnidaria),big.mark = ",")
cnidaria_max <- format(max(summary_statistics$Cnidaria),big.mark = ",")
cnidaria_median <- format(median(summary_statistics$Cnidaria),big.mark = ",")

#number of Porifera annotations
porifera_min <- format(min(summary_statistics$Porifera),big.mark = ",")
porifera_max <- format(max(summary_statistics$Porifera),big.mark = ",")
porifera_median <- format(median(summary_statistics$Porifera),big.mark = ",")

#number of Echinodermata annotations
echinodermata_min <- format(min(summary_statistics$Echinodermata),big.mark = ",")
echinodermata_max <- format(max(summary_statistics$Echinodermata),big.mark = ",")
echinodermata_median <- format(median(summary_statistics$Echinodermata),big.mark = ",")

#number of Vertebrata annotations
vertebrata_min <- format(min(summary_statistics$Vertebrata),big.mark = ",")
vertebrata_max <- format(max(summary_statistics$Vertebrata),big.mark = ",")
vertebrata_median <- format(median(summary_statistics$Vertebrata),big.mark = ",")

#calculate min/max/median for percent unidentified at each taxonomic level
annotations_taxonomy_percentage_wide <- summary_statistics |> 
  mutate(across(class:species, \(x) x/phylum*100, .names = "{.col}_percent")) |> 
  select("dive_number", class = "class_percent", order = "order_percent", family = "family_percent", genus = "genus_percent", species = "species_percent")

annotations_taxonomy_percentage_wide[is.na(annotations_taxonomy_percentage_wide)] = 0

class_min <- round(min(annotations_taxonomy_percentage_wide$class),0)
class_max <- round(max(annotations_taxonomy_percentage_wide$class),0)
class_median <- round(median(annotations_taxonomy_percentage_wide$class),0)

order_min <- round(min(annotations_taxonomy_percentage_wide$order),0)
order_max <- round(max(annotations_taxonomy_percentage_wide$order),0)
order_median <- round(median(annotations_taxonomy_percentage_wide$order),0)

family_min <- round(min(annotations_taxonomy_percentage_wide$family),0)
family_max <- round(max(annotations_taxonomy_percentage_wide$family),0)
family_median <- round(median(annotations_taxonomy_percentage_wide$family),0)

genus_min <- round(min(annotations_taxonomy_percentage_wide$genus),0)
genus_max <- round(max(annotations_taxonomy_percentage_wide$genus),0)
genus_median <- round(median(annotations_taxonomy_percentage_wide$genus),0)

species_min <- round(min(annotations_taxonomy_percentage_wide$species),0)
species_max <- round(max(annotations_taxonomy_percentage_wide$species),0)
species_median <- round(median(annotations_taxonomy_percentage_wide$species),0)
```

# Data Summary

This report is an expedition-level summary of expert annotations of benthic deep-sea communities and substrate for groups of remotely operated vehicle (ROV) dives within the NOAA Ocean Exploration [Atlantic Seafloor Partnership for Integrated Research and Exploration (ASPIRE)](https://oceanexplorer.noaa.gov/explorations/aspire/welcome.html) campaign (2017-2022). Associated NOAA Ocean Exploration cruise reports are available from the [NOAA Institutional Repository](https://repository.library.noaa.gov/). Video was collected using protocols described in the [NOAA Ocean Exploration ROV and Telepresence Deepwater Exploration Procedures manual](https://doi.org/10.25923/n605-za83).

These annotations have undergone quality control by the France Lab at the University of Louisiana Lafayette following protocols described in the [Annotating NOAA Ocean Exploration Video guidance](https://oceanexplorer.noaa.gov/okeanos/collaboration-tools/science_annotations/media/NOAA_Ocean_Exploration_Video_Annotations.pdf).[^1] These summarized annotations only represent the annotations made during the benthic portion of the dives (defined using the on- and off-bottom times in the Dive Summary Reports available for each dive on the [expedition-specific landing page](https://www.ncei.noaa.gov/waf/okeanos-rov-cruises/)) and do not include any annotations that may have been made during ROV descent/ascent or during water column exploration. This report also does not contain any annotations made during maritime heritage dives.

[^1]: The annotations in SeaTube can be continuously updated by users. Annotations may have been modified or updated since the quality control process by the France Lab.

This report provides data visualizations in the form of tables and graphs of:

-   Dive summary metrics (ROV time on bottom; ROV mean depth),

-   Annotation metrics by dive (total numbers of benthic biological and geoform/substrate annotations; percentage of annotations identified to each taxonomic level; phylum-level community composition; numbers of annotations of taxonomic groups of interest, i.e., Cnidaria, Porifera, Echinodermata, and Vertebrata), and

-   Quality control metrics (percentage of annotations flagged for review at the time of report generation; number of unidentified biological annotations).

# Annotation Information

Input data are .csv files sourced from the [Ocean Networks Canada SeaTube website](https://data.oceannetworks.ca/SeaTubeSearch).

This report was generated using annotations downloaded from SeaTube on `r annotations_date`.

# Expedition Summary

`r expedition_name` consisted of `r total_dive_count` total dives, of which `r UCH_dives` explored maritime heritage sites and the remaining `r science_dive_count` dives explored the benthos and water column. ROV time on bottom ranged from `r bottom_min` to `r bottom_max` hours with a median of `r bottom_median` hours across all dives. The shallowest dive had a mean depth of `r depth_min` meters, the deepest dive had a mean depth of `r depth_max` meters, and the median across all dives was `r depth_median` meters. The median number of geoform or substrate annotations was `r geo_sub_annotations_median`, and median number of benthic biological annotations was `r bio_annotations_median`.

# Dive Summary Metrics

```{r}
#| echo: false
#| label: ROV-time-on-bottom

ggplot(summary_statistics, aes(x = dive_number, y = bottom_time_hours)) +
  geom_col(fill = "#0085CA") +
  labs(x = "Dive Number", y = "Hours") +
  scale_x_continuous(n.breaks = nrow(summary_statistics)) +
  coord_cartesian(xlim = c(1, nrow(summary_statistics))) +
  OER_theme
```

###### [Figure 1]{custom-style="Strong"}. ROV time on bottom in hours.

ROV time on bottom ranged from `r bottom_min` hours to `r bottom_max` hours during `r expedition_name` with a median time on bottom of `r bottom_median` hours.

```{r}
#| echo: false
#| 
#this is on hold until we decide on a strategy to obtain this metric for 
#pre-2020 expeditions


#if (summary_statistics$benthic_start[1] > "2020-01-01"){
#gplot(summary_statistics, aes(x = dive_number, y = distance_traveled_m)) +
  #geom_col(fill = "#0085CA") +
  #labs(x = "Dive Number", y = "Meters") +
  #scale_x_continuous(n.breaks = nrow(summary_statistics)) +
  #OER_theme

#distance_min <- min(summary_statistics$distance_traveled_m)
#distance_max <- max(summary_statistics$distance_traveled_m)
#distance_median <- median(summary_statistics$distance_traveled_m)
#}
```

```{r}
#| echo: false

ggplot(summary_statistics, aes(x = dive_number, y = mean_depth)) +
  geom_col(fill = "#0085CA") +
  labs(x = "Dive Number", y = "Depth (m)") +
  scale_x_continuous(n.breaks = nrow(summary_statistics)) +
  coord_cartesian(xlim = c(1, nrow(summary_statistics))) +
  OER_theme
```

###### [Figure 2]{custom-style="Strong"}. Mean ROV depth in meters during the benthic portion of each dive.

The shallowest dive during `r expedition_name` had a mean depth of `r depth_min`, and the deepest dive had a mean depth of `r depth_max`. Overall, dives during `r expedition_name` had a median depth of `r depth_median` meters.

# Annotation Metrics

```{r}
#| echo: false

ggplot(summary_statistics, aes(x = dive_number, y = phylum)) +
  geom_col(fill = "#0085CA") +
  labs(x = "Dive Number", y = "Number of Annotations") +
  scale_x_continuous(n.breaks = nrow(summary_statistics)) +
  coord_cartesian(xlim = c(1, nrow(summary_statistics))) +
  OER_theme
```

###### [Figure 3]{custom-style="Strong"}. Total biological annotations.

The total number of benthic biological annotations (all phyla) for each dive during `r expedition_name` ranged from `r bio_annotations_min` to `r bio_annotations_max` with a median of `r bio_annotations_median` annotations across all dives.

```{r}
#| echo: false

# annotations_taxonomy <- summary_statistics |> 
#   select("dive_number",total = "phylum","class","order","family","genus","species") |> 
#   pivot_longer(total:species) |> 
#   rename(taxonomic_level = name,
#          annotation_count = value) |> 
#   mutate(taxonomic_level = as.factor(taxonomic_level),
#          dive_number = as.factor(dive_number)) |> 
#   mutate(taxonomic_level = fct_relevel(taxonomic_level,
#                                        c("total","class","order",
#                                          "family","genus","species")))
#View(annotations_taxonomy)

# ggplot(annotations_taxonomy, aes(fill = taxonomic_level, 
#                                  y = annotation_count,
#                                  x = dive_number)) +
#   geom_col(width = 1, position = "dodge2") +
#   scale_fill_manual(values = c('#001489', '#153099', '#1d4aa8', '#1f64b8', '#1a7ec8', '#0099d8')) +
#   #coord_flip() +
#   labs(title = "Biological Annotations by Taxonomic Level", 
#        x = "Dive Number", y = "Number of Annotations", fill = "Taxonomic Level") +
#   OER_theme +
#   theme(legend.position = "bottom",
#         legend.title = element_text(family = "Calibri", color = "#003087", size = 10),
#         legend.text = element_text(family = "Calibri", color = "#003087", size = 10))
```

```{r}
#| echo: false
ggplot(annotations_taxonomy_percentage, aes(x = taxonomic_level,
                                 y = annotation_percentage,
                                 group = dive_number,
                                 fill = taxonomic_level)) +
         geom_col(show.legend = FALSE) +
         facet_wrap(~ dive_number) +
  labs(x = "Dive Number", y = "Percentage of Total Annotations") +
  scale_fill_manual(values = c('#003087', '#1051a1', '#1b74b9', '#2698ce', '#34bddf')) +
  theme_bw() + 
  theme(plot.title = element_text(family = "Roboto", color = "#003087", size = 16),
        axis.title = element_text(family = "Roboto", color = "#003087", size = 12),
        axis.text.y = element_text(family = "Roboto", size = 8),
        axis.text.x = element_text(family = "Roboto", size = 8, angle = -35, hjust = 0),
        panel.grid.major.y = element_line(color = "gray50"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        plot.margin = margin(0.5,0.5,0.5,1.5, "cm"))
```

###### [Figure 4]{custom-style="Strong"}. Percentage of annotations identified to each taxonomic level

One common problem in deep-sea biology is that many taxa observed in video can only be identified to high taxonomic levels, like phylum or class, and cannot be identified to lower taxonomic levels, such as genus or species. The figure above shows the percentage of biological annotations (all phyla) that could be identified to each taxonomic level, from class through species, for each dive during `r expedition_name`.

Across all dives:

-   The percentage of benthic biological annotations that could be identified to the class level ranged from `r class_min`-`r class_max` with a median of `r class_median`%.

-   The percentage of benthic biological annotations that could be identified to the order level ranged from `r order_min`-`r order_max` with a median of `r order_median`%.

-   The percentage of benthic biological annotations that could be identified to the family level ranged from `r family_min`-`r family_max` with a median of `r family_median`%.

-   The percentage of benthic biological annotations that could be identified to the genus level ranged from `r genus_min`-`r genus_max` with a median of `r genus_median`%.

-   The percentage of benthic biological annotations that could be identified to the species level ranged from `r species_min`-`r species_max` with a median of `r species_median`%.

```{r}
#| echo: false
phyla_frequency_percent_all <- phyla_frequency_percent_all |> 
  mutate(dive_number = as.factor(dive_number)) |>
  arrange(phylum)

ggplot(phyla_frequency_percent_all, aes(fill = percent, x = dive_number, y = phylum)) +
  geom_tile(color = "gray", aes(width = 0.85, height = 0.999)) +
  scale_fill_gradient(low = "white", high = "#003087", guide = "legend", name = "Percent") +
  theme(panel.border = element_rect(fill = NA, color = alpha("black", 0.5))) +
  labs(x = "Dive Number") +
  OER_theme +
  theme(legend.title = element_text(color = "#003087", family = "Roboto", size = 12), legend.text = element_text(family = "Roboto", size = 8), panel.grid = element_blank(), axis.title.y = element_blank())

```

###### [Figure 5]{custom-style="Strong"}. Phylum-level annotations and community composition

Phyla annotated during `r expedition_name` varied by dive and are illustrated here using color intensity, which indicates the relative contribution of each phylum to the total benthic biological annotations for each dive. The most common phyla annotated during EX2104 can be identified by examining the variation in color intensity across rows. The community composition annotated during EX2104 can be compared across dives by examining the variation in color intensity across columns.

```{r}
#| echo: false

ggplot(summary_statistics, aes(x = dive_number, y = Cnidaria)) +
  geom_col(fill = "#0085CA") +
  labs(x = "Dive Number", y = "Number of Annotations") +
  scale_x_continuous(n.breaks = nrow(summary_statistics)) +
  coord_cartesian(xlim = c(1, nrow(summary_statistics))) +
  OER_theme
```

###### [Figure 6]{custom-style="Strong"}. Total Cnidaria annotations.

Cnidaria is a phylum of particular interest as it contains a variety of habitat-building invertebrates commonly observed in the deep sea, including corals and sea anemones. The total number of cnidarians annotated during each dive of `r expedition_name` ranged from `r cnidaria_min` to `r cnidaria_max` with a median of `r cnidaria_median` cnidarian annotations across all dives.

```{r}
#| echo: false

ggplot(summary_statistics, aes(x = dive_number, y = Porifera)) +
  geom_col(fill = "#0085CA") +
  labs(x = "Dive Number", y = "Number of Annotations") +
  scale_x_continuous(n.breaks = nrow(summary_statistics)) +
  coord_cartesian(xlim = c(1, nrow(summary_statistics))) +
  OER_theme
```

###### [Figure 7]{custom-style="Strong"}. Total Porifera annotations.

The phylum Porifera includes sponges. The total number of porifera annotated during each dive of `r expedition_name` ranged from `r porifera_min` to `r porifera_max` with a median of `r porifera_median` porifera annotations across all dives.

```{r}
#| echo: false

ggplot(summary_statistics, aes(x = dive_number, y = Echinodermata)) +
  geom_col(fill = "#0085CA") +
  labs(x = "Dive Number", y = "Number of Annotations") +
  scale_x_continuous(n.breaks = nrow(summary_statistics)) +
  coord_cartesian(xlim = c(1, nrow(summary_statistics))) +
  OER_theme
```

###### [Figure 8]{custom-style="Strong"}. Total Echinodermata annotations.

The phylum Echinodermata includes a variety of invertebrates common to the deep sea, including sea stars (Class: Asteroidea), brittle stars (Class: Ophiuroidea), crinoids (Class: Crinoidea), sea urchins (Class: Echinoidea), and sea cucumbers (Class: Holothuroidea). The total number of echinoderms annotated during each dive of `r expedition_name` ranged from `r echinodermata_min` to `r echinodermata_max` with a median of `r echinodermata_median` echinoderm annotations across all dives.

```{r}
#| echo: false

ggplot(summary_statistics, aes(x = dive_number, y = Vertebrata)) +
  geom_col(fill = "#0085CA") +
  labs(x = "Dive Number", y = "Number of Annotations") +
  scale_x_continuous(n.breaks = nrow(summary_statistics)) +
  coord_cartesian(xlim = c(1, nrow(summary_statistics))) +
  OER_theme
```

###### [Figure 9]{custom-style="Strong"}. Total Vertebrata annotations.

The subphylum Vertebrata contains all vertebrates within the phylum Chordata. Commonly observed vertebrates include fishes. The total number of vertebrates annotated during each dive of `r expedition_name` ranged from `r vertebrata_min` to `r vertebrata_max` with a median of `r vertebrata_median` vertebrate annotations across all dives.

```{r}
#| echo: false

ggplot(summary_statistics, aes(x = dive_number, y = geoform_or_substrate)) +
  geom_col(fill = "#0085CA") +
  labs(x = "Dive Number", y = "Number of Annotations") +
  scale_x_continuous(n.breaks = nrow(summary_statistics)) +
  coord_cartesian(xlim = c(1, nrow(summary_statistics))) +
  OER_theme
```

###### [Figure 10]{custom-style="Strong"}. Total geoform or substrate annotations.

The number of geoform and substrate annotations within each dive during `r expedition_name` ranged from `r geo_sub_annotations_min` to `r geo_sub_annotations_max` with a median of `r geo_sub_annotations_median` annotations across all dives.

# Quality control metrics

```{r}
#| echo: false

percent_flagged_rounded <- round(summary_statistics$percent_flagged,1)

ggplot(summary_statistics, aes(x = dive_number, y = percent_flagged)) +
  geom_col(fill = "#0085CA") +
  labs(x = "Dive Number", y = "Percentage of Total Annotations") +
  ylim(0,100) +
  scale_x_continuous(n.breaks = nrow(summary_statistics)) +
  coord_cartesian(xlim = c(1, nrow(summary_statistics))) +
  geom_text(aes(label = percent_flagged_rounded), vjust = -0.2, family = "Roboto", color = "#0085CA", size = 3.5) + 
  OER_theme
```

###### [Figure 11]{custom-style="Strong"}. Percentage of annotations flagged for review.

The percentage of annotations flagged for review for each dive of `r expedition_name` ranged from `r percent_flagged_min` to `r percent_flagged_max` with a median of `r percent_flagged_median` annotations flagged for review across all dives.

```{r}
#| echo: false

ggplot(summary_statistics, aes(x = dive_number, y = Unidentified_Biota)) +
  geom_col(fill = "#0085CA") +
  labs(x = "Dive Number", y = "Number of Annotations") +
  scale_x_continuous(n.breaks = nrow(summary_statistics)) +
  coord_cartesian(xlim = c(1, nrow(summary_statistics))) +
  OER_theme

```

###### [Figure 12]{custom-style="Strong"}. Number of unidentified biological annotations.

The number of benthic biological annotations that were unable to be identified to the phylum level during `r expedition_name` ranged from `r unID_bio_min` to `r unID_bio_max` with a median of `r unID_bio_median` unidentified biological annotations across all dives.
