---
title: "EX2104: 2021 North Atlantic Stepping Stones"
format:
  docx: 
    reference-doc: 2025_OER_for-quarto.docx
editor: visual
---

```{r}
#| echo: false
#| output: false

#set expedition information to use throughout document
expedition_name <- "EX2104"
expedition_title <- "North Atlantic Stepping Stones"
annotations_date <- "10-21-2024"
UCH_dives <- 0

#libraries
library(readr)
library(ggplot2)
library(showtext)
library(tidyr)
library(dplyr)
library(forcats)

#add Roboto to showtext package
font_add_google("Roboto", "Roboto")
showtext_auto()

#create OER figure theme to apply throughout the report
OER_theme <- theme_bw() + theme(plot.title = element_text(family = "Roboto", color = "#003087", size = 16),
        axis.title = element_text(family = "Roboto", color = "#003087", size = 12),
        axis.text = element_text(family = "Roboto", size = 10))

#load summary statistics
summary_statistics <- read_csv(paste0("summary_statistics_",expedition_name,".csv"),col_names = TRUE, show_col_types = FALSE)
#load phylum frequency percent
phyla_frequency_percent_all <- read_csv(paste0("phyla_frequency_percent_all_",expedition_name,".csv"),col_names = TRUE)

#calculate total dive number
science_dive_count <- nrow(summary_statistics)
total_dive_count <- science_dive_count + UCH_dives

#calculate percentage of unidentified annotations at each taxonomic level
annotations_taxonomy_percentage <- summary_statistics |> 
  mutate(across(class:species, \(x) x/phylum*100, .names = "{.col}_percent")) |> 
  select("dive_number", class = "class_percent", order = "order_percent", family = "family_percent", genus = "genus_percent", species = "species_percent") |> 
  pivot_longer(class:species) |> 
  rename(taxonomic_level = name,
         annotation_percentage = value) |> 
  mutate(taxonomic_level = as.factor(taxonomic_level),
         dive_number = as.factor(dive_number)) |> 
  mutate(taxonomic_level = fct_relevel(taxonomic_level,
                                       c("class","order",
                                         "family","genus","species")))
annotations_taxonomy_percentage[is.na(annotations_taxonomy_percentage)] = 0

#perform min/max/median calculations across dives for each data type

#ROV bottom time
bottom_min <- round(min(summary_statistics$bottom_time_hours),2)
bottom_max <- round(max(summary_statistics$bottom_time_hours),2)
bottom_median <- round(median(summary_statistics$bottom_time_hours),2)

#Mean depth
depth_min <- round(min(summary_statistics$mean_depth),0)
depth_max <- round(max(summary_statistics$mean_depth),0)
depth_median <- round(median(summary_statistics$mean_depth),0)

#number of geoform/substrate annotations
geo_sub_annotations_min <- min(summary_statistics$geoform_or_substrate)
geo_sub_annotations_max <- max(summary_statistics$geoform_or_substrate)
geo_sub_annotations_median <- median(summary_statistics$geoform_or_substrate)

#number of biological annotations
bio_annotations_min <- min(summary_statistics$phylum)
bio_annotations_max <- max(summary_statistics$phylum)
bio_annotations_median <- median(summary_statistics$phylum)

#percentage annotations flagged for review
percent_flagged_min <- round(min(summary_statistics$percent_flagged),1)
percent_flagged_max <- round(max(summary_statistics$percent_flagged),1)
percent_flagged_median <- round(median(summary_statistics$percent_flagged),1)

#number of unidentified biological annotations
unID_bio_min <- min(summary_statistics$Unidentified_Biota)
unID_bio_max <- max(summary_statistics$Unidentified_Biota)
unID_bio_median <- median(summary_statistics$Unidentified_Biota)

#number of Cnidaria annotations
cnidaria_min <- min(summary_statistics$Cnidaria)
cnidaria_max <- max(summary_statistics$Cnidaria)
cnidaria_median <- median(summary_statistics$Cnidaria)

#number of Porifera annotations
porifera_min <- min(summary_statistics$Porifera)
porifera_max <- max(summary_statistics$Porifera)
porifera_median <- median(summary_statistics$Porifera)

#number of Echinodermata annotations
echinodermata_min <- min(summary_statistics$Echinodermata)
echinodermata_max <- max(summary_statistics$Echinodermata)
echinodermata_median <- median(summary_statistics$Echinodermata)

#number of Vertebrata annotations
vertebrata_min <- min(summary_statistics$Vertebrata)
vertebrata_max <- max(summary_statistics$Vertebrata)
vertebrata_median <- median(summary_statistics$Vertebrata)

#calculate min/max/median for percent unidentified at each taxonomic level
annotations_taxonomy_percentage_wide <- summary_statistics |> 
  mutate(across(class:species, \(x) x/phylum*100, .names = "{.col}_percent")) |> 
  select("dive_number", class = "class_percent", order = "order_percent", family = "family_percent", genus = "genus_percent", species = "species_percent")

annotations_taxonomy_percentage_wide[is.na(annotations_taxonomy_percentage_wide)] = 0

class_min <- round(min(annotations_taxonomy_percentage_wide$class),0)
class_max <- round(max(annotations_taxonomy_percentage_wide$class),0)
class_median <- round(median(annotations_taxonomy_percentage_wide$class),0)

order_min <- round(min(annotations_taxonomy_percentage_wide$order),0)
order_max <- round(max(annotations_taxonomy_percentage_wide$order),0)
order_median <- round(median(annotations_taxonomy_percentage_wide$order),0)

family_min <- round(min(annotations_taxonomy_percentage_wide$family),0)
family_max <- round(max(annotations_taxonomy_percentage_wide$family),0)
family_median <- round(median(annotations_taxonomy_percentage_wide$family),0)

genus_min <- round(min(annotations_taxonomy_percentage_wide$genus),0)
genus_max <- round(max(annotations_taxonomy_percentage_wide$genus),0)
genus_median <- round(median(annotations_taxonomy_percentage_wide$genus),0)

species_min <- round(min(annotations_taxonomy_percentage_wide$species),0)
species_max <- round(max(annotations_taxonomy_percentage_wide$species),0)
species_median <- round(median(annotations_taxonomy_percentage_wide$species),0)
```

# Data Summary

This report is an expedition-level summary of expert annotations of benthic deep-sea communities and substrate for groups of ROV dives within the NOAA Ocean Exploration [ASPIRE](https://oceanexplorer.noaa.gov/explorations/aspire/welcome.html) campaign (2017-2022). Video was collected using protocols described in the NOAA Ocean Exploration ROV and Telepresence Deepwater Exploration Procedures [manual](https://doi.org/10.25923/n605-za83). These data have been QAQC\'d by the France Lab at the University of Louisiana Lafayette, following protocols described in the Annotating NOAA Ocean Exploration Video [manual](https://oceanexplorer.noaa.gov/okeanos/collaboration-tools/science_annotations/media/NOAA_Ocean_Exploration_Video_Annotations.pdf). These summarized annotations only represent the annotations made during the benthic portion of the dive (defined using the on- and off-bottom times provided in the [Dive Summary .txt files](https://www.ncei.noaa.gov/waf/okeanos-rov-cruises/) available for each dive) and do not include any annotations that may have been made during ROV descent/ascent or within midwater transects. This report also does not contain any annotations made during UCH dives, which do not have available Dive Summary .txt files.

This report provides data visualizations in the form of tables and graphs of: 1. Dive summary metrics (ROV time on bottom; ROV mean depth); 2. Observational metrics by dive (total numbers of benthic biological and geoform/substrate annotations; percentage of annotations identified to each taxonomic level; phylum-level community composition; numbers of annotations of taxonomic groups of interest, including Cnidaria, Porifera, Echinodermata, and Vertebrata); and 3. Quality control metrics (percentage of annotations flagged for review; number of unidentified biological annotations).

# Annotation Information

Input data are .csv files sourced from the [Ocean Networks Canada SeaTube website](https://data.oceannetworks.ca/SeaTubeSearch).

This report was generated using annotations downloaded from SeaTube on `r annotations_date`.

# Expedition Summary

`r expedition_name` consisted of `r total_dive_count` total dives, of which `r science_dive_count` were science-focused and `r UCH_dives` explored underwater cultural heritage sites.

# Dive summary metrics

```{r}
#| echo: false
#| label: ROV-time-on-bottom

ggplot(summary_statistics, aes(x = dive_number, y = bottom_time_hours)) +
  geom_col(fill = "#0085CA") +
  labs(x = "Dive Number", y = "Hours") +
  scale_x_continuous(n.breaks = nrow(summary_statistics)) +
  OER_theme
```

###### [Figure 1]{custom-style="Strong"}. ROV time on bottom in hours.

ROV time on bottom ranged from `r bottom_min` hours to `r bottom_max` hours during `r expedition_name`, with a median time on bottom of `r bottom_median` hours.

```{r}
#| echo: false
#| 
#this is on hold until we decide on a strategy to obtain this metric for 
#pre-2020 expeditions


#if (summary_statistics$benthic_start[1] > "2020-01-01"){
#gplot(summary_statistics, aes(x = dive_number, y = distance_traveled_m)) +
  #geom_col(fill = "#0085CA") +
  #labs(x = "Dive Number", y = "Meters") +
  #scale_x_continuous(n.breaks = nrow(summary_statistics)) +
  #OER_theme

#distance_min <- min(summary_statistics$distance_traveled_m)
#distance_max <- max(summary_statistics$distance_traveled_m)
#distance_median <- median(summary_statistics$distance_traveled_m)
#}
```

```{r}
#| echo: false

ggplot(summary_statistics, aes(x = dive_number, y = mean_depth)) +
  geom_col(fill = "#0085CA") +
  labs(x = "Dive Number", y = "Depth (m)") +
  scale_x_continuous(n.breaks = nrow(summary_statistics)) +
  OER_theme
```

###### [Figure 2]{custom-style="Strong"}. Mean ROV depth in meters during the benthic portion of each dive.

The shallowest dive during `r expedition_name` had a mean depth of `r depth_min`, and the deepest dive had a mean depth of `r depth_max`. Overall, dives within `r expedition_name` had a median depth of `r depth_median` meters.

# Observational metrics

```{r}
#| echo: false

ggplot(summary_statistics, aes(x = dive_number, y = geoform_or_substrate)) +
  geom_col(fill = "#34bddf") +
  labs(x = "Dive Number", y = "Number of Annotations") +
  scale_x_continuous(n.breaks = nrow(summary_statistics)) +
  OER_theme
```

###### [Figure 3]{custom-style="Strong"}. Total geoform or substrate annotations.

The number of geoform and substrate annotations within each dive during `r expedition_name` ranged from `r geo_sub_annotations_min` to `r geo_sub_annotations_max` with a median of `r geo_sub_annotations_median` annotations .

```{r}
#| echo: false

ggplot(summary_statistics, aes(x = dive_number, y = phylum)) +
  geom_col(fill = "#34bddf") +
  labs(x = "Dive Number", y = "Number of Annotations") +
  scale_x_continuous(n.breaks = nrow(summary_statistics)) +
  OER_theme
```

###### [Figure 4]{custom-style="Strong"}. Total biological annotations.

The total number of benthic biological annotations within each dive during `r expedition_name` ranged from `r bio_annotations_min` to `r bio_annotations_max` with a median of `r bio_annotations_median` across all dives.

```{r}
#| echo: false

# annotations_taxonomy <- summary_statistics |> 
#   select("dive_number",total = "phylum","class","order","family","genus","species") |> 
#   pivot_longer(total:species) |> 
#   rename(taxonomic_level = name,
#          annotation_count = value) |> 
#   mutate(taxonomic_level = as.factor(taxonomic_level),
#          dive_number = as.factor(dive_number)) |> 
#   mutate(taxonomic_level = fct_relevel(taxonomic_level,
#                                        c("total","class","order",
#                                          "family","genus","species")))
#View(annotations_taxonomy)

# ggplot(annotations_taxonomy, aes(fill = taxonomic_level, 
#                                  y = annotation_count,
#                                  x = dive_number)) +
#   geom_col(width = 1, position = "dodge2") +
#   scale_fill_manual(values = c('#001489', '#153099', '#1d4aa8', '#1f64b8', '#1a7ec8', '#0099d8')) +
#   #coord_flip() +
#   labs(title = "Biological Annotations by Taxonomic Level", 
#        x = "Dive Number", y = "Number of Annotations", fill = "Taxonomic Level") +
#   OER_theme +
#   theme(legend.position = "bottom",
#         legend.title = element_text(family = "Calibri", color = "#003087", size = 10),
#         legend.text = element_text(family = "Calibri", color = "#003087", size = 10))
```

```{r}
#| echo: false
ggplot(annotations_taxonomy_percentage, aes(x = taxonomic_level,
                                 y = annotation_percentage,
                                 group = dive_number,
                                 fill = taxonomic_level)) +
         geom_col(show.legend = FALSE) +
         facet_wrap(~ dive_number) +
  labs(x = "Dive Number", y = "Percentage of Total Annotations") +
  scale_fill_manual(values = c('#003087', '#1051a1', '#1b74b9', '#2698ce', '#34bddf')) +
  theme_bw() + 
  theme(plot.title = element_text(family = "Roboto", color = "#003087", size = 16),
        axis.title = element_text(family = "Roboto", color = "#003087", size = 12),
        axis.text.y = element_text(family = "Roboto", size = 8),
        axis.text.x = element_text(family = "Roboto", size = 8, angle = -35, hjust = 0),
        panel.grid.major.y = element_line(color = "gray50"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        plot.margin = margin(0.5,0.5,0.5,1.5, "cm"))
```

###### [Figure 5]{custom-style="Strong"}. Percentage of annotations identified to each taxonomic level

One common problem in deep-sea biology is that many taxa observed in video can only be identified to high taxonomic levels, like phylum or class, and are not able to be identified to lower taxonomic levels, such as genus or species. The figure above shows the percentage of biological annotations that could be identified to each taxonomic level, from class through species, within each dive within `r expedition_name`.

Across all dives, the median percentage of benthic biological annotations that could be identified to the class level was `r class_median`%, with the range `r class_min` - `r class_max`. The median percentage of benthic biological annotations that could be identified to the order level was `r order_median`%, range `r order_min` - `r order_max`. The median percentage of benthic biological annotations that could be identified to the family level was `r family_median`%, range `r family_min` - `r family_max`. The median percentage of benthic biological annotations that could be identified to the genus level was `r genus_median`%, range `r genus_min` - `r genus_max`. The median percentage of benthic biological annotations that could be identified to the species level was `r species_median`%, range `r species_min` - `r species_max`.

```{r}
#| echo: false
phyla_frequency_percent_all <- phyla_frequency_percent_all |> 
  mutate(dive_number = as.factor(dive_number)) |>
  arrange(phylum)

ggplot(phyla_frequency_percent_all, aes(fill = percent, x = dive_number, y = phylum)) +
  geom_tile(color = "gray", aes(width = 0.85, height = 0.999)) +
  scale_fill_gradient(low = "white", high = "#003087", guide = "legend", name = "Percent") +
  theme(panel.border = element_rect(fill = NA, color = alpha("black", 0.5))) +
  labs(x = "Dive Number") +
  OER_theme +
  theme(legend.title = element_text(color = "#003087", family = "Roboto", size = 12), legend.text = element_text(family = "Roboto", size = 8), panel.grid = element_blank(), axis.title.y = element_blank())

```

###### [Figure 6]{custom-style="Strong"}. Phylum-level community composition

The figure above shows the phyla observed across all dives within `r expedition_name`. The color intensity indicates the relative contribution of each phylum to the total benthic biological annotations within each dive. The most common phyla observed can be identified by comparing the variation in color intensity across rows. The community composition can be compared across dives by observing the variation in color intensity across columns.

```{r}
#| echo: false

ggplot(summary_statistics, aes(x = dive_number, y = Cnidaria)) +
  geom_col(fill = "#34bddf") +
  labs(x = "Dive Number", y = "Number of Annotations") +
  scale_x_continuous(n.breaks = nrow(summary_statistics)) +
  OER_theme
```

###### [Figure 7]{custom-style="Strong"}. Total Cnidaria annotations.

Cnidaria is a phylum of particular interest as it contains a variety of habitat-building invertebrates commonly observed in the deep sea, including corals and sea anemones. The total number of cnidarians observed across dives within `r expedition_name` ranged from `r cnidaria_min` to `r cnidaria_max`, with a median of `r cnidaria_median`.

```{r}
#| echo: false

ggplot(summary_statistics, aes(x = dive_number, y = Porifera)) +
  geom_col(fill = "#34bddf") +
  labs(x = "Dive Number", y = "Number of Annotations") +
  scale_x_continuous(n.breaks = nrow(summary_statistics)) +
  OER_theme
```

###### [Figure 8]{custom-style="Strong"}. Total Porifera annotations.

The phylum Porifera includes sponges. The total number of porifera observations across dives within `r expedition_name` ranged from `r porifera_min` to `r porifera_max`, with a median of `r porifera_median`.

```{r}
#| echo: false

ggplot(summary_statistics, aes(x = dive_number, y = Echinodermata)) +
  geom_col(fill = "#34bddf") +
  labs(x = "Dive Number", y = "Number of Annotations") +
  scale_x_continuous(n.breaks = nrow(summary_statistics)) +
  OER_theme
```

###### [Figure 9]{custom-style="Strong"}. Total Echinodermata Annotations.

The phylum Echinodermata includes a variety of invertebrates common to the deep sea, including seastars (Class: Asteroidea), brittle stars (Class: Ophiuroidea), crinoids (Class: Crinoidea), sea urchins (Class: Echinoidea), and sea cucumbers (Class: Holothuroidea). The total number of echinoderms observed across dives within `r expedition_name` ranged from `r echinodermata_min` to `r echinodermata_max`, with a median of `r echinodermata_median`.

```{r}
#| echo: false

ggplot(summary_statistics, aes(x = dive_number, y = Vertebrata)) +
  geom_col(fill = "#34bddf") +
  labs(x = "Dive Number", y = "Number of Annotations") +
  scale_x_continuous(n.breaks = nrow(summary_statistics)) +
  OER_theme
```

###### [Figure 10]{custom-style="Strong"}. Total Vertebrata Annotations.

The subphylum Vertebrata contains all vertebrates within the phylum Chordata. Commonly observed vertebrates include fishes. The total number of vertebrates observed across dives within `r expedition_name` ranged from `r vertebrata_min` to `r vertebrata_max`, with a median of `r vertebrata_median`.

# Quality control metrics

```{r}
#| echo: false

percent_flagged_rounded <- round(summary_statistics$percent_flagged,1)

ggplot(summary_statistics, aes(x = dive_number, y = percent_flagged)) +
  geom_col(fill = "#003087") +
  labs(x = "Dive Number", y = "Percentage of Total Annotations") +
  ylim(0,100) +
  scale_x_continuous(n.breaks = nrow(summary_statistics)) +
  geom_text(aes(label = percent_flagged_rounded), vjust = -0.2, family = "Roboto", color = "#003087") + 
  OER_theme
```

###### [Figure 11]{custom-style="Strong"}. Percentage of annotations flagged for review.

The percentage of annotations flagged for review within each dive during `r expedition_name` ranged from `r percent_flagged_min` to `r percent_flagged_max` with a median of `r percent_flagged_median` across all dives.

```{r}
#| echo: false

ggplot(summary_statistics, aes(x = dive_number, y = Unidentified_Biota)) +
  geom_col(fill = "#003087") +
  labs(x = "Dive Number", y = "Number of Annotations") +
  scale_x_continuous(n.breaks = nrow(summary_statistics)) +
  OER_theme

```

###### [Figure 12]{custom-style="Strong"}. Number of unidentified biological annotations.

The number of benthic biological annotations that were unable to be identified to the phylum level during `r expedition_name` varied from `r unID_bio_min` to `r unID_bio_max` with a median of `r unID_bio_median` across all dives.
